Revisiting Short-Chain Fatty Acids and the Microbiota in Colorectal Cancer

### 2018-05-22 submission/manuscript.Rmd
update methods and add first results section

@@

@@

**Study design and sampling.** The overall protocol has been described in detail previously [@normalization_sze2017; @crc_model_baxter2016]. In brief, this study used fecal samples obtained at either a single cross-sectional time point (n=`r length(metai$sample)`) or from before and after treatment for their tumor (n=`r length(metaf$initial)`). For patients undergoing treatment for their tumor the length of time between their initial and follow up sample ranged from `r min(metaf$time)` - `r max(metaf$time)` days. Our use of treatment has been previously defined as encompassing removal of a tumor with or without chemotherapy and radiation [@normalization_sze2017]. Diagnosis of tumor was made by colonoscopic examination and histopathological review of biopsies obtained [@normalization_sze2017; @crc_model_baxter2016]. The University of Michigan Institutional Review Board approved the study and informed consent was obtained from all participants in accordance to the guidelines set out by the Helsinki Decleration.

@@

**Obtaining OPFs from metagenomes.** A subset of the cross-sectional group (n=`r length(metai$sample)`) containing a total of `r length(geof_samples$X1)` individuals (normal n=`r filter(geof_samples, X30 == "Healthy") %>% pull(X30) %>% length()` normal, adenoma n=`r filter(geof_samples, X30 == "Adenoma") %>% pull(X30) %>% length()`, and carcinoma n=`r filter(geof_samples, X30 == "Cancer") %>% pull(X30) %>% length()`) was shotgun sequenced on an Illumina HiSeq with 125 bp paired end reads using a previously described method [@Hannigan2017]. Briefly, the sequences were quality filtered and sequences aligning to the human genome were removed prior to contig assembly with MEGAHIT [@Li2015]. Open Reading Frames (ORFs) were identified using Prodigal [@Hyatt2010], counts generated using Diamond [@Buchfink2014], subsequent clustering into Operational Protein Families (OPFs) used mmseq2 [@Steinegger2017], and OPF gene alignment used the KEGG database [@Kanehisa2015].

@@

**Statistical analysis workflow.** All analysis was performed using the statistical language R [@r_citation_2017]. Generally, differences between the different disease groups used a Kruskal-Walis rank sum test with a Dunn's post-hoc test. We next assessed whether SCFAs added to Random Forest models built with OTU data significantly improved classification. Next, models to classify high or low SCFA concentration based on 16S rRNA gene sequencing data were created using Random Forest [@randomforest_citation_2002]. Regression models to classify the exact SCFA concentration based on 16S rRNA gene sequencing data also were built using the Random Forest algorithm. The measured SCFA concentrations were first tested for differences between groups. The ability of 16S rRNA gene sequencing to classify these concentrations were then assessed. Next, the imputed gene counts of important mediators of SCFA creation were tested. Finally, the counts generated for OPFs that matched important genes involved with SCFA creation were analyzed.

>>>>

# Read in analysis of SCFA by HPLC
hplc_groups_scfa_kruskal <- read_csv("../data/process/tables/acetate_kruskal_crc_groups.csv") %>% 
  mutate(scfa = "acetate") %>% slice(1) %>% 
  bind_rows(
    read_csv("../data/process/tables/butyrate_kruskal_crc_groups.csv") %>% 
      mutate(scfa = "butyrate") %>% slice(1), 
    read_csv("../data/process/tables/propionate_kruskal_crc_groups.csv") %>% 
      mutate(scfa = "propionate") %>% slice(1)) %>% 
  mutate(bh = p.adjust(pvalue, method = "BH"))

hplc_treat_scfa_kruskal <- read_csv("../data/process/tables/scfa_treatment_comparison_pvalues.csv") %>% 
  filter(scfa != "isobutyrate")

# Load in Random Forest with and without SCFA summary data
disease_rf <- read_csv("../data/process/tables/rf_model_ttest_comparisons.csv")

@@

**Decreased SCFA concentrations are not associated with tumors.** We used frozen fecal samples from `r length(metai$sample)` individuals and HPLC to measure acetate, butyrate, and propionate concentrations at a cross-sectional point in time. There was no difference between individuals with normal colons and patients with either an adenoma or carcinoma for any of the SCFAs measured after multiple comparison correction (P-value `r ifelse(min(pull(hplc_groups_scfa_kruskal, bh)) > 0.15, invisible(">"), invisible("<"))` 0.15) [Figure 1A - 1C]. We next measured the concentration of SCFAs in `r length(metaf$initial)` patients with an adenoma or carcinoma in which we had pre- and post-treatment fecal samples. Although there was a general trend for increasing acetate, butyrate, and propionate concentrations after treatment for tumors, there was no significant difference pre- and post-treatment for patients with adenoma (P-value `r ifelse(min(filter(hplc_treat_scfa_kruskal, disease == "adenoma") %>% pull(bh)) > 0.20, invisible(">"), invisible("<"))` 0.20) or carcinoma (P-value `r ifelse(min(filter(hplc_treat_scfa_kruskal, disease == "carcinoma") %>% pull(bh)) > 0.80, invisible(">"), invisible("<"))` 0.80) [Figure 1D - 1F]. Even though there was no change in SCFA concentrations between individuals with normal colons and those with tumors, this information could still be important to help classify disease.

**Random Forest models with SCFA concentrations do not classify tumor better.**

@@

**Study design and sampling.** The overall protocol has been described in detail previously [@normalization_sze2017; @crc_model_baxter2016]. In brief, this study used fecal samples obtained at either a single cross-sectional time point (n=`r length(metai$sample)`) or from before (pre-) and after (post-) treatment for their tumor (n=`r length(metaf$initial)`). For patients undergoing treatment for their tumor the length of time between their initial and follow up sample ranged from `r min(metaf$time)` - `r max(metaf$time)` days. Our use of treatment has been previously defined as encompassing removal of a tumor with or without chemotherapy and radiation [@normalization_sze2017]. Diagnosis of tumor was made by colonoscopic examination and histopathological review of biopsies obtained [@normalization_sze2017; @crc_model_baxter2016]. The University of Michigan Institutional Review Board approved the study and informed consent was obtained from all participants in accordance to the guidelines set out by the Helsinki Decleration.

@@

**Obtaining OPFs from metagenomes.** A subset of the cross-sectional group (n=`r length(metai$sample)`) containing a total of `r length(geof_samples$X1)` individuals (normal n=`r filter(geof_samples, X30 == "Healthy") %>% pull(X30) %>% length()` normal, adenoma n=`r filter(geof_samples, X30 == "Adenoma") %>% pull(X30) %>% length()`, and carcinoma n=`r filter(geof_samples, X30 == "Cancer") %>% pull(X30) %>% length()`) was shotgun sequenced on an Illumina HiSeq using 125 bp paired end reads and a previously described method [@Hannigan2017]. Briefly, the sequences were quality filtered and sequences aligning to the human genome were removed prior to contig assembly with MEGAHIT [@Li2015]. Open Reading Frames (ORFs) were identified using Prodigal [@Hyatt2010], counts generated using Diamond [@Buchfink2014], subsequent clustering into Operational Protein Families (OPFs) used mmseq2 [@Steinegger2017], and OPF alignment used the KEGG database [@Kanehisa2015].

@@

**Statistical analysis workflow.** All analysis was performed using the statistical language R [@r_citation_2017]. Generally, differences between the different disease groups used a Kruskal-Walis rank sum test with a Dunn's post-hoc test. We next assessed whether SCFAs added to Random Forest models built with OTU data significantly improved classification. Next, models to classify high or low SCFA concentration based on 16S rRNA gene sequencing data were created using Random Forest [@randomforest_citation_2002]. Regression models to classify the exact SCFA concentration based on 16S rRNA gene sequencing data also were built using the Random Forest algorithm. The measured SCFA concentrations were first tested for differences between groups. The ability of 16S rRNA gene sequencing to classify these concentrations were then assessed. Next, the imputed gene counts of important mediators of SCFA creation were tested. Finally, the counts generated for OPFs that matched important genes involved with SCFA creation were analyzed. Where appropriate Benjamini-Hochberg was used to correct for multiple comparisons [@benjamini_controlling_1995.

### 2018-05-22 submission/manuscript.tex
update methods and add first results section

@@

\textbf{Study design and sampling.} The overall protocol has been
described in detail previously (1, 2). In brief, this study used fecal
samples obtained at either a single cross-sectional time point (n=490)
or from before and after treatment for their tumor (n=67). For patients
undergoing treatment for their tumor the length of time between their
initial and follow up sample ranged from 188 - 546 days. Our use of
treatment has been previously defined as encompassing removal of a tumor
with or without chemotherapy and radiation (1). Diagnosis of tumor was
made by colonoscopic examination and histopathological review of
biopsies obtained (1, 2). The University of Michigan Institutional
Review Board approved the study and informed consent was obtained from
all participants in accordance to the guidelines set out by the Helsinki
Decleration.

@@

\textbf{Obtaining OPFs from metagenomes.} A subset of the
cross-sectional group (n=490) containing a total of 85 individuals
(normal n=29 normal, adenoma n=28, and carcinoma n=28) was shotgun
sequenced on an Illumina HiSeq with 125 bp paired end reads using a
previously described method (8). Briefly, the sequences were quality
filtered and sequences aligning to the human genome were removed prior
to contig assembly with MEGAHIT (9). Open Reading Frames (ORFs) were
identified using Prodigal (10), counts generated using Diamond (11),
subsequent clustering into Operational Protein Families (OPFs) used
mmseq2 (12), and OPF gene alignment used the KEGG database (13).

@@

\textbf{Statistical analysis workflow.} All analysis was performed using
the statistical language R (15). Generally, differences between the
different disease groups used a Kruskal-Walis rank sum test with a
Dunn's post-hoc test. We next assessed whether SCFAs added to Random
Forest models built with OTU data significantly improved classification.
Next, models to classify high or low SCFA concentration based on 16S
rRNA gene sequencing data were created using Random Forest (16).
Regression models to classify the exact SCFA concentration based on 16S
rRNA gene sequencing data also were built using the Random Forest
algorithm. The measured SCFA concentrations were first tested for
differences between groups. The ability of 16S rRNA gene sequencing to
classify these concentrations were then assessed. Next, the imputed gene
counts of important mediators of SCFA creation were tested. Finally, the
counts generated for OPFs that matched important genes involved with
SCFA creation were analyzed.

@@

>>>>

\textbf{Decreased SCFA concentrations are not associated with tumors.}
There was no difference between individuals with normal colons and
patients with either an adenoma or carcinoma after multiple comparison
correction for any of the SCFAs measured (P-value \textgreater{} 0.15)
{[}Figure 1A - 1C{]}. Although there was a trend for increasing acetate,
butyrate, and propionate concentrations after treatment for tumors,
there was no significant difference pre- and post-treatment for
individuals with adenoma (P-value \textgreater{} 0.20) or carcinoma
(P-value \textgreater{} 0.80) {[}Figure 1D - 1F{]}.

@@

\textbf{Study design and sampling.} The overall protocol has been
described in detail previously (1, 2). In brief, this study used fecal
samples obtained at either a single cross-sectional time point (n=490)
or from before (pre-) and after (post-) treatment for their tumor
(n=67). For patients undergoing treatment for their tumor the length of
time between their initial and follow up sample ranged from 188 - 546
days. Our use of treatment has been previously defined as encompassing
removal of a tumor with or without chemotherapy and radiation (1).
Diagnosis of tumor was made by colonoscopic examination and
histopathological review of biopsies obtained (1, 2). The University of
Michigan Institutional Review Board approved the study and informed
consent was obtained from all participants in accordance to the
guidelines set out by the Helsinki Decleration.

@@

\textbf{Obtaining OPFs from metagenomes.} A subset of the
cross-sectional group (n=490) containing a total of 85 individuals
(normal n=29 normal, adenoma n=28, and carcinoma n=28) was shotgun
sequenced on an Illumina HiSeq using 125 bp paired end reads and a
previously described method (8). Briefly, the sequences were quality
filtered and sequences aligning to the human genome were removed prior
to contig assembly with MEGAHIT (9). Open Reading Frames (ORFs) were
identified using Prodigal (10), counts generated using Diamond (11),
subsequent clustering into Operational Protein Families (OPFs) used
mmseq2 (12), and OPF alignment used the KEGG database (13).

@@

\textbf{Statistical analysis workflow.} All analysis was performed using
the statistical language R (15). Generally, differences between the
different disease groups used a Kruskal-Walis rank sum test with a
Dunn's post-hoc test. We next assessed whether SCFAs added to Random
Forest models built with OTU data significantly improved classification.
Next, models to classify high or low SCFA concentration based on 16S
rRNA gene sequencing data were created using Random Forest (16).
Regression models to classify the exact SCFA concentration based on 16S
rRNA gene sequencing data also were built using the Random Forest
algorithm. The measured SCFA concentrations were first tested for
differences between groups. The ability of 16S rRNA gene sequencing to
classify these concentrations were then assessed. Next, the imputed gene
counts of important mediators of SCFA creation were tested. Finally, the
counts generated for OPFs that matched important genes involved with
SCFA creation were analyzed. Where appropriate Benjamini-Hochberg was
used to correct for multiple comparisons {[}(17).

@@

\hypertarget{ref-benjamini_controlling_1995}{}
17. \textbf{Benjamini Y}, \textbf{Hochberg Y}. 1995. Controlling the
false discovery rate: A practical and powerful approach to multiple
testing. Journal of the Royal Statistical Society Series B
(Methodological) \textbf{57}:289--300.
